---
title: "Identifying Screening-Relevant Context in an OSA Study Using Clinical Note Metadata and LLM-Extracted Signals"
subtitle: ""
author: "Ashley Batugo"
date: today
format:
  html:
    css: "osa-report-theme.css"
    toc: true
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    page-layout: full
    anchor-sections: true
include-in-header: |
  <link href="https://fonts.googleapis.com/css2?family=Playfair+Display:wght@700;900&family=Source+Sans+Pro:wght@300;400;600&display=swap" rel="stylesheet">
---

## Overview

This project examined which note-level contextual and metadata features are associated with clinical research coordinator (CRC) exclusion decisions during screening for an NIH-funded Obstructive Sleep Apnea (OSA) study. CRCs rely on two categories of information found in unstructured EHR notes: true clinical contraindications such as active medical instability and operational skip signals including recent surgery, hospitalization or pending procedures. Using LLM extracted note-level evidence, I constructed a de-identified, note level analytic dataset and used logistic regression as an exploratory modeling approach to identify which metadata features were most associated with both informal and formal exclusionary contexts. Discussions with Dr. Danielle Mowery and Emily Schriver helped shape the dataset design and modeling strategy and Paula Salvador, the CRC for this study, provided insights into the pre-screening process and the reasons behind choosing not to reach out to certain patients. The materials for this project can be found in this [Github repository](https://github.com/ashbatu/BMIN503_Final_Project).

## Introduction {#sec-introduction}

Clinical research is essential for advancing medical knowledge, particularly for conditions that are often underrecognized and underdiagnosed, such as OSA ([Motamedi et al., 2009](https://pmc.ncbi.nlm.nih.gov/articles/PMC3096276/)). Similar to clinical trials, prospective clinical studies depend on many factors including strong study design, careful planning, timely recruitment, and sustained participation retention ([Lai et al., 2019](https://journals.sagepub.com/doi/full/10.1177/1740774519829709?casa_token=mm71iIUcfGkAAAAA%3AvRwJV3xTHnNLQ2-WsNICmuzN-IQLbLRioaR-lmo3lrqEWKBfZ-WFaB21-bRNLq34dCeeWsCGKWi9Yg)). However, identifying eligible patients remains one of the major challenges in clinical research ([Cai et al., 2021](https://acrjournals.onlinelibrary.wiley.com/doi/10.1002/acr2.11289#:~:text=The%20LiiRA%20study%20team%20incorporated,not%20screening%20out%20eligible%20patients.)). Although researchers increasingly rely on electronic health records (EHRs) to support recruitment, determining whether a patient should be contacted still requires detailed manual review of unstructured EHR notes. This process is time-consuming, requires clinical judgement, and often incorporates not only formal contraindications but also operational 'skip signals', which affect whether initiating contact is appropriate. Because of this, chart review can occupy several hours of CRC time each day, slowing recruitment and adding substantial operational burden ([Etchberger, 2016](https://www.linkedin.com/pulse/chart-review-should-sponsors-pay-clinical-research-sites-etchberger#:~:text=Some%20trials%20remain%20very%20complicated,takes%20to%20find%20those%20patients.)). These challenges are particularly critical for milestone-driven NIH-funded projects, where delays in meeting recruitment goals can jeopardize continued funding. This project was motivated by an ongoing NIH-funded OSA clinical study in which our team has faced recruitment delays due to how resource-intensive the chart review process is for the CRCs.

Addressing this problem requires collaboration with experts from different fields including medicine, informatics, and clinical research operations. Clinicians provide the judgement needed to determine which patients should be contacted, assist with recruitment within their own patient populations, and interpret the context of the notes and patient charts. Informatics and data science contribute the methods to extract, organize, and analyze unstructured EHR data to help more efficiently determine whether is it suitable to contact patients for recruit. In developing this project, conversations with Dr. Danielle Mowery, Director of the Institute for Biomedical Informatics (CIC), and Emily Schriver, a translational data scientist in the CIC, helped clarify how informatics can be applied to identify meaningful EHR features that support clinical teams in improving recruitment workflows. This problem is also closely tied to clinical research operations, since improving recruitment directly benefits those responsible for identifying, reaching out to, and enrolling participants.

## Methods {#sec-methods}

#### Methods Overview

This project involved two methodological phrases.

**Phase 1** focused on extracting note-level evidence related to the exclusion criteria using a Large Language Model (LLM) to classify unstructured notes in the EHR within a secure, HIPAA-compliant environment. This phase included identifying the study population, retrieving and de-identifying relevant notes, developing and freezing the LLM prompt, performing a light face-validity check on a small sample, and applying the prompt to assign each note to any of the three exclusion-context buckets.

**Phase 2** used the de-identified, LLM-labeled notes to construct the analytic dataset ad evaluate which combinations of EHR note metadata (specialty, note type, encounter type, and temporal window) most effectively reveal clinically relevant contexts that influence CRC outreach decisions . This phase involved categorical feature engineering, handling missing metadata, and fitting logistic regression models in R to examine both the overall exclusion-context signals and category-specific patterns. All R data for feature engineering and modeling are included in this report.

#### PHASE 1 - LLM-Based Evidence Extraction

::: callout-important
No functional code is included for Phase 1 because this phrase involved protected health information (PHI) and was completed entirely within Penn Medicine's HIPAA-compliant environments (Databricks and the LPC cluster). All patient identification, note retrieval, de-identification and LLM processing were completed using SQL, Python, and R inside these secure workspaces. Important code snippets used during this phase are included.
:::

[**Study Population**]{.underline}

Patients were identified from a CRC-maintained aggregated spreadsheet containing all individuals who had recent visits a Penn Medicine Sleep Center and were automatically and manually reviewed during screening for the NIH-funded OSA clinical trial. For this project, patients were included if:

1.  They underwent manual chart review by the CRC, and
2.  They were not recruited due to an exclusion classified as "Medical Condition" or "Other"

From this group, the cohort were further restricted to patients with non-administrative exclusion signals that are most likely captured within unstructured clinical, pathology, and surgical notes (e.g. cancer, panic disorder, recent surgery, non-OSA sleep condition such as narcolepsy) which require extensive manual review.

\[SHOW CODE SNIPPET\]

[**Data Sources and Note Retrieval**]{.underline}

Clinical notes were extracted from the Epic Clarity database on the Penn Medicine Azure Databricks environment. We included clinical notes (e.g. progress notes, discharge summaries, ED notes), and all surgical and pathology notes to capture both operational skip signals and true contraindications. Note-level metadata (note type, encounter specialty, and encounter type) were also retrieved. All notes within one year prior to the CRC's pre-screening date were included, and each note was assigned to one of four temporal windows (0--30, 31--90, 91--180, and \>180 days).

\[SHOW CODE FOR TEMPORAL WINDOWS\]

Additionally, each note was prefixed with a standardized header indicating the temporal context of the note relative to the pre-screening date: `[TIME_RELATIVE_TO_PRESCREEN: <WINDOW_BIN> | DELTA_DAYS = <NUMBER>]`.

\[SHOW CODE FOR PREFIX\]

Empty notes and notes deemed sensitive by the Penn Medicine Privacy Office were removed.

\[SHOE PYTHON CODE FOR REMOVING NA'S\]

The remaining notes were then deidentified using an adapted PHIlter implementation on the LPC cluster.

[**Exclusion Category Bucketing**]{.underline}

Exclusion reasons were consolidated into higher level exclusion buckets due to the sparsity of the individual exclusion signals after reviewing the exclusion notes assigned by the CRC. The buckets were then used to organize how the LLM identified exclusion signals in the notes. Exclusions were grouped into the following three buckets

| Exclusion Bucket                                 | Description                                                                                                                                     |
|------------------------|------------------------------------------------|
| Clinical Contraindications (`clinical_contra`)   | Major current clinical conditions that influence the decision to reach out to a patient because of medical instability of likely ineligibility. |
| Procedural & Recent Events (`procedural_recent`) | Recent, ongoing, or upcoming procedures or clinical events that indicate current acute clinical episodes or need for recovery.                  |
| Sleep-Specific Conditions (`sleep_specific`)     | Sleep-related diagnoses that indicate a patient has a non-OSA sleep disorder or condition.                                                      |

\[SHOW CODE TO CONSOLIDATE EXCLUSIONS\]

[**LLM Prompt Development, Evaluation, and Note-Level Output**]{.underline}

The GPT-4o mini chat model available in the Penn Medicine Azure Databricks environment was used to evaluate each note and assign a 0/1 decision for each of the three exclusion categories. The prompt instructed the LLM to read each note (with its temporal prefix) and determine, for each category, whether the note contained information meeting that category's exclusion criteria (1 = meets criteria, 0 = does not), along with a brief rationale and a confidence score.

The prompt consisted of the following components:

-   a brief description of the LLM's role and the overall classification task,

-   clarification of the temporal prefix added to each note,

-   definitions and examples of the exclusion categories and the constraints for assigning a 1 or 0,

-   the required output elements for each category: a binary assignment, a short rationale, and a confidence score, and

-   the standardized output format.

The full prompt sent to the LLM is included in the Appendix.

Abridged Python and SQL code used in the Databricks environment were included below to illustrate how notes were submitted to the LLM and how resulting outputs were parsed for downstream use.

The GPT-4o mini endpoint was invoked programatically and the LLM ran the prompt as shown below:

\[INCLUDE FUNCTION TO CALL THE PROMPT\]

An example of the formatted note input sent to the LLM is shown here (synthetic):

\[SHOW EXAMPLE HERE -- SYNTHETIC VERSION\]

The model returned a structured JSON response containing three sets of binary labels, rationales, and confidence scores. A representative synthetic JSON output is shown below:

\[SHOW EXAMPLE HERE - SYNTHETIC VERSION -- JSON\]

This JSON was then parsed into a dataframe for creating the analytical dataset:

\[SHOW PYTHON CODE TO PARSE THE JSON\]

To assess the quality of the prompt, patients were split in 60/40 training and testing sets while also maintaining the proportion of the rolled up exclusion buckets. The prompt was first applied to all notes for patients in the training set. Performance of the prompt was evaluated based on (1) how well the LLM identified the exclusion categories and (2) a manual review of 5 patients per category to ensure that the LLM-generated rationales aligned with the note text. Once ≥80% coverage was achieved, the prompt was frozen and then applied to the entire dataset (training and testing notes). Coverage was computed with the following code:

\[CODE TO EVALUATE COVERAGE\]

#### PHASE 2 - Regression Modeling and Interpretation of Note Metadata Predictors

[**FEATURE ENGINEERING**]{.underline}

**Loading Required Packages**

Preprocessing the analytical dataset and regression model was conducted using the following packages:

```{r}
require(tidyverse) 
require(modelsummary)
require(DescTools)
```

**Loading the De-Identified Note-Level Dataset**

The de-identified LLM results created in Phase 1 were exported from Databricks as a CSV and imported as a dataframe into R. Each row represents a single clinical note with its associated metadata and LLM-derived exclusion predictions. The data was loaded as follows:

```{r}
notes_metadata <- read_csv('../datasets/notes_data.csv')
notes_metadata
```

**Handling Null Values and Converting Metadata Fields to Factor Type**

Because Databricks exports missing character fields as the literal string 'null', these must be converted to proper NA values:

```{r}
notes_metadata <- notes_metadata %>% 
  mutate(across(where(is.character), ~na_if(., "null")))
notes_metadata
```

To prevent rows from being dropped during `glm()` due to missing factor levels, missing values were recoded as "Unknown" and the core note metadata fields were converted to factors:

```{r}
notes_metadata <- notes_metadata %>% 
  mutate(across(c(ip_note_type, note_type, specialty,
                  encounter_type, window_bin, source), 
                ~ case_when(is.na(.x) ~ 'Unknown',
                          T ~ .x))) %>% 
  mutate(across(c(ip_note_type, note_type, specialty,
                  encounter_type, window_bin, source), 
                ~ as.factor(.x))) 
notes_metadata 
```

**Creating Modeling Outcomes**

In addition to the three note-level exclusion categories (`clinical_contra`, `procedural_recent`, and `sleep_specific`) that were created in Phase 1, a new combined "overall exclusion" was created. This new variable identifies whether any of the exclusion signals (clinical, procedural, or sleep-specific) was present in a note. This new outcome allows for conducting a more general analysis of note contexts that are associated with exclusion relevant signals regardless of the category.

```{r}
notes_metadata <- notes_metadata %>% 
  mutate(pred_overall_excl = 
           case_when(pred_clinical_contra == 1 | 
                       pred_procedural_recent == 1 | 
                       pred_sleep_specific == 1 ~ 1,
                     T ~ 0)) %>% 
  relocate(pred_overall_excl, .after = pred_sleep_specific)
notes_metadata
```

**Inspecting Metadata Distributions Prior to Collapsing**

Since this dataset is a relatively small sample of notes (n = 1,735 notes), I checked to see the count of notes associated with each factor level to assess how smaller sample sizes can be collapsed and joined with other levels. This was done because sparse categories can destabilize logistic regression leading to overfitting, biased estimates, and overestimating odds ratios.

```{r}
table(notes_metadata$ip_note_type)
table(notes_metadata$note_type)
table(notes_metadata$specialty)
table(notes_metadata$encounter_type)
table(notes_metadata$window_bin)

# Factor Collapsing
notes_metadata <- notes_metadata %>% 
  mutate(
    ip_note_type = case_when(
      ip_note_type %in% c("Brief Op Note", "Op Note", "OR PostOp", "OR PreOp") ~ "Operative Note",
      ip_note_type %in% c("H&P", "Interval H&P Note") ~ "H&P Note",
      ip_note_type %in% c("ED Notes", "ED Provider Notes") ~ "ED Note",
      ip_note_type == "Progress Notes" ~ "Progress Note",
      ip_note_type == "Discharge Summary" ~ "Discharge Summary",
      ip_note_type == "Unknown" ~ "Unknown",
      TRUE ~ ip_note_type
    ),
    ip_note_type = factor(ip_note_type),

    note_type = case_when(
      note_type == "SURGICAL PATHOLOGY REPORT" ~ "Pathology Report",
      TRUE ~ note_type
    ),

    specialty = str_trim(specialty),
    specialty = if_else(is.na(specialty) | specialty == "", "Unknown", specialty),

    specialty = case_when(
      specialty %in% c("GYN", "Gynecology", "OB/Gyn") ~ "OB/Gyn",
      specialty %in% c("GI", "GIS", "Gastroenterology", "GI Surgery") ~ "Gastroenterology/GI",
      specialty %in% c("ORTHO", "Orthopaedics", "Orthopedics") ~ "Orthopedics",
      specialty %in% c("Urology", "UROLOGY") ~ "Urology",
      specialty %in% c('PMR', "Physical Medicine and Rehab") ~ 
        'PM&R',
      specialty %in% c('Hematology/Oncology', 'Oncology') ~ 'Heme/Onc',
      specialty %in% c("ORL", "Otorhinolaryngology") ~ "ENT",
      specialty %in% c(
        "Colon and Rectal Surgery",
        "Oral Maxillofacial Surgery",
        "Plastic Surgery",
        "PLASSURG",
        "Thoracic Surgery",
        "THORSURG",
        "Transplant Surgery",
        "VASCSURG",
        "Vascular Surgery",
        "EOS",
        "CRS"
      ) ~ "Surgery - Other",
      TRUE ~ specialty
    ),

    specialty = fct_lump_min(factor(specialty), min = 10, other_level = "Other Specialty"),
    specialty = factor(specialty),
    encounter_type = str_trim(encounter_type),

    encounter_type = case_when(
      encounter_type == "Office Visit" ~ "Office Visit",
      encounter_type == "Telemedicine" ~ "Telemedicine",

      encounter_type %in% c("Hospital Encounter",
                            "Post Hospitalization",
                            "Post Emergency") ~ "Hospital Encounter",

      encounter_type %in% c("Procedure", "Procedure Visit",
                            "Infusion Visit", "Medication Management",
                            "Orders Only") ~ "Procedure/Treatment",

      encounter_type == "Reconciled Outside Data" ~ "Reconciled Outside Data",

      # Administrative / communication / scheduling
      encounter_type %in% c(
        "Care Management", "Allied Health Visit", "Allied Health Visit (Non-Chargeable)",
        "Appointment", "Letter (Out)", "Telephone", "Out of Office Visit",
        "Patient Outreach", "Transitions in Care", "No Show-No Charge",
        "Psych Care Management", "Social Work (Non-Chargeable)",
        "Enrollment", "CCBH Scheduled", "CCBH Unscheduled"
      ) ~ "Ancillary Encounter",

      encounter_type %in% c("Unknown", "Research",
                            "Research Encounter", "Research (Non-Chargeable)") ~ "Other Encounter",

      TRUE ~ "Other Encounter"
    ),

    encounter_type = factor(encounter_type)
  ) 
notes_metadata
```

**Harmonizing Note Type Fields**

In the dataset there are two fields that capture the note type: `ip_note_type` and `note_type`. As shown in the code below, these fields tend to overlap. To address this, I create one harmonized field called `note_type_final`. This field is selects the non-Unknown value when one field is populated, collapsed when both fields have the same value, or Unknown when neither field is populated:

```{r}
notes_metadata %>% 
  distinct(ip_note_type, note_type)
```

```{r}
notes_metadata <- notes_metadata %>% 
  mutate(note_type_final = 
           factor(case_when(ip_note_type != 'Unknown' & note_type == 'Unknown' ~ ip_note_type,
                     ip_note_type == "Unknown" & note_type != 'Unknown' ~ note_type,
                     ip_note_type == "Unknown" & note_type == "Unknown" ~ 'Unknown',
                     ip_note_type == note_type ~ note_type))) %>% 
  relocate(note_type_final, .after = note_type)
notes_metadata
```

After pre-processing and collapsing factors, I performed final quality checks of the analytic dataset before modeling by:

-   assessing the distribution of notes based on each exclusion prediction and

```{r}
table(notes_metadata$pred_overall_excl)
table(notes_metadata$pred_clinical_contra)
table(notes_metadata$pred_procedural_recent)
table(notes_metadata$pred_sleep_specific)
```

Because the counts for the positive class for `pred_procedural_recent` and `pred_sleep_specific` are low and to replicate the CRCs pre-screening process, I created a new prediction category called `pred_other_excl` (in her pre-screeing process she categorizes exclusions by 'Medical Condition' and 'Other'). 'Medical Condition' exclusions will be represented through `pred_procedural_recent` and the 'Other' excluions will be represtented through `pred_other_excl` (the combination of `pred_procedural_recent` and `pred_sleep_specific`)

```{r}
notes_metadata <- notes_metadata %>% 
  mutate(pred_other_excl = 
           case_when(pred_procedural_recent == 1 | 
                       pred_sleep_specific == 1 ~ 1,
                     T ~ 0)) %>% 
  relocate(pred_other_excl, .after = pred_overall_excl)
notes_metadata
```

New distribution with `pred_other_excl` field:

```{r}
table(notes_metadata$pred_other_excl)
```

-   ensuring that factors have been properly collapsed for each metadata fields

```{r}
table(notes_metadata$note_type_final)   
table(notes_metadata$specialty)
table(notes_metadata$encounter_type)
table(notes_metadata$window_bin)
```

Because logistic regression is used for exploratory purposes, I did address class imbalances.

**Building the final modeling dataset**

Finally, I created the model-ready dataset for regression. This includes the four outcome variables (`pred_overall_excl`, `pred_clinical_contra`, `pred_procedural_recent`, and `pred_sleep_specific`) and the note metadata predictors (`note_type_final`, `specialty`, `encounter_type`, and `window_bin`). A row identifier was also added to the dataset and `source` which indicates the type of note (clinical, pathology, or surgical note) for descriptive statistics and characterizing the dataset:

```{r}
model_df <- notes_metadata %>% 
  select(pred_overall_excl, pred_clinical_contra, 
         pred_procedural_recent, pred_sleep_specific,
         pred_other_excl,
         note_type_final, 
         specialty,
         encounter_type, 
         window_bin,
         source) %>% 
  mutate(row = row_number())
model_df
```

**Modeling Strategy**

To identify the most relevant note metadata features that are strongly associated with exclusion-relevant notes, logistic regression models were used. Logistic regression was selected because it identifies the influence of predictors simultaneously and the strength of association among the predictors and the outcome.

For this project, three models were fit:

1.  an overall exclusion model to identify the most influential metadata features across all of the exclusion buckets (outcome variable: `pred_overall_excl`)
2.  category-specific models for clinical (Medical conditions), and other (recent procedural, and sleep-specific) exclusions.

Because this project is exploratory, models were fit with the entire dataset to maximize statistical power and the certainty of the parameter estimates. Bootstrapping was then conducted to evaluate whether the feature effects remained consistent with repeated sampling.

Describe the data used and general methodological approach used to address the problem described in the @sec-introduction. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why.

**TODO**

-   Talk about data cleaning --\> grouping this was talked about in phrase 2

-   Justify the predictor variables and the predicted variables

-   Why did you pick the covariates --\> make sure that this is clearly explained --\> talk about the features specifically that are being used for the models

## Results {#sec-results}

### Descriptive Summaries

```{r}
require(kableExtra)
# total counts
overall_counts <- notes_metadata %>%
  summarize(
    total_notes = n_distinct(note_id),
    total_patients = n_distinct(pat_id),
    total_encounters = n_distinct(pat_enc_csn_id),
    
  )

# Notes per patient 
notes_per_patient <- notes_metadata %>%
  group_by(pat_id) %>%
  summarise(n_notes = n())

notes_per_patient_summary <- notes_per_patient %>%
  summarise(
    median_notes = median(n_notes),
    IQR_notes = IQR(n_notes),
    min_notes = min(n_notes),
    max_notes = max(n_notes)
  )

# Min and Max Note Dates
note_dates_min_max <- notes_metadata %>% 
  mutate(min_note = min(note_service_dttm),
         max_note = max(note_service_dttm)) %>% 
  distinct(min_note, max_note)

dataset_overview_table <- bind_cols(overall_counts, notes_per_patient_summary, note_dates_min_max) %>% 
  mutate(note_date_range = paste0(format(min_note, "%m/%d/%Y"), ' to ', format(max_note, "%m/%d/%Y"))) %>% 
  mutate(`Notes Per Patient` = paste0("Median: ", median_notes, ", (Min: ", min_notes, ", Max: ", max_notes, ")")) %>%
  select(`Total Notes` = total_notes, `Total Patients` = total_patients, `Total Encounters` = total_encounters, `Notes Per Patient`, `Note Date Range` = note_date_range)
  
tibble::tibble(
  `Cohort characteristic` = c(
    "Total notes",
    "Unique patients",
    "Unique encounters",
    "Notes per patient",
    "Note Date Range"
  ),
  `Overall` = c(
    dataset_overview_table$`Total Notes`,
    dataset_overview_table$`Total Patients`,
    dataset_overview_table$`Total Encounters`,
    dataset_overview_table$`Notes Per Patient`,
    dataset_overview_table$`Note Date Range`
  )
) %>% kable(
    caption = "Table 1. Study cohort and note characteristics",
    align = "l"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

```{r}
library(gtsummary)

table_2_notelevelmetadata <- model_df %>%
  select(
    note_type_final,
    specialty,
    encounter_type,
    window_bin,
    source
  ) %>%
  tbl_summary(        
    percent  = "column",
    missing  = "no",
    statistic = all_categorical() ~ "{n} ({p}%)"
  ) %>%
  modify_header(
    label ~ "**Note-level metadata**"
  ) %>%
  modify_caption(
    "**Table 2. Distribution of notes by note-level metadata**"
  ) %>%
  bold_labels()

table_2_notelevelmetadata
```

```{r}
# standard styling
theme_osa <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title      = element_text(face = "bold", size = 15),
      axis.title      = element_text(face = "bold"),
      axis.text.x     = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.text       = element_text(color = "gray20"),
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_line(color = "#E5E7EB"),
      axis.line       = element_line(color = "#111827"),
      legend.position = "top",
      legend.title    = element_text(face = "bold"),
      legend.text     = element_text(size = 12),
      # plot.margin     = margin(12, 14, 10, 14),
      panel.background = element_rect(fill = "white", color = NA)
    )
}

osa_palette <- c(
  "#2C7FB8",  # muted blue
  "#7FCDBB",  # soft teal
  "#EDF8B1",  # pale yellow-green
  "#FEC44F",  # warm amber
  "#FC9272",  # blush coral
  "#9ECAE1",  # light slate blue
  "#A1D99B",  # mint
  "#BCBDDC"   # soft lavender
)
# expanding color palette above
osa_palette_expanded <- colorRampPalette(osa_palette)(50)
```

```{r}
# distribution of notes by note type
note_type_plot <- model_df %>% 
  ggplot(aes(x = fct_infreq(note_type_final), fill = note_type_final)) +
  
  geom_bar(width = 0.75) +
  
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    vjust = -0.4,
    size = 3,
    color = "#1F2937"   # dark slate
  ) +
  
  scale_fill_manual(values = osa_palette) +
  
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.08))
  ) +
  
  labs(
    title = "Distribution of Clinical Notes by Note Type",
    x = "Note type",
    y = "Number of notes",
    fill = "Note type"
  ) +
  
  theme_osa() +
  
  theme(
    legend.position = "none"   # remove legend (redundant)
  )
note_type_plot

# ggsave(
#   "../figs/note_type_plot.png",
#   plot = note_type_plot,
#   width = 5.5,   # wide
#   height = 3.0,  # short
#   dpi = 300
# )
```


```{r}
# distribution of specialties
specialties_plot <- model_df %>% 
  ggplot(aes(x = fct_infreq(specialty), fill = specialty)) +
  
  geom_bar(width = 0.75) +
  
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    vjust = -0.4,
    size = 3,
    color = "#1F2937"   # dark slate
  ) +
  
  scale_fill_manual(values = osa_palette_expanded) +
  
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.08))
  ) +
  
  labs(
    title = "Distribution of Clinical Notes by Specialty",
    x = "Specialty",
    y = "Number of notes",
    fill = "Specialty"
  ) +
  
  theme_osa() +
  
  theme(
    legend.position = "none"   # remove legend (redundant)
  )
specialties_plot

# ggsave(
#   "../figs/specialties_plot.png",
#   plot = specialties_plot,
#   width = 5.5,   # wide
#   height = 3.0,  # short
#   dpi = 300
# )
```

```{r}
# distribution of encounter type
encounter_type_plots <- model_df %>% 
  ggplot(aes(x = fct_infreq(encounter_type), fill = encounter_type)) +
  
  geom_bar(width = 0.75) +
  
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    vjust = -0.4,
    size = 3,
    color = "#1F2937"   # dark slate
  ) +
  
  scale_fill_manual(values = osa_palette_expanded) +
  
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.08))
  ) +
  
  labs(
    title = "Distribution of Clinical Notes by Encounter Type",
    x = "Encounter Type",
    y = "Number of notes",
    fill = "Encounter Type"
  ) +
  
  theme_osa() +
  
  theme(
    legend.position = "none"   # remove legend (redundant)
  )
encounter_type_plots

# ggsave(
#   "../figs/encounter_type_plots.png",
#   plot = encounter_type_plots,
#   width = 5.5,   # wide
#   height = 3.0,  # short
#   dpi = 300
# )
```

```{r}
# distribution of time windows 
time_windows_plot <- model_df %>% 
  ggplot(aes(x = fct_infreq(window_bin), fill = window_bin)) +
  
  geom_bar(width = 0.75) +
  
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    vjust = -0.4,
    size = 3,
    color = "#1F2937"   # dark slate
  ) +
  
  scale_fill_manual(values = osa_palette) +
  
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.08))
  ) +
  
  labs(
    title = "Distribution of Clinical Notes by Time Window",
    x = "Time Window Bin",
    y = "Number of notes",
    fill = "Time Window Bin"
  ) +
  
  theme_osa() +
  
  theme(
    legend.position = "none"   # remove legend (redundant)
  )
time_windows_plot

# ggsave(
#   "../figs/time_windows_plot.png",
#   plot = time_windows_plot,
#   width = 5.5,   # wide
#   height = 3.0,  # short
#   dpi = 300
# )
```

A total of 2,911 notes for 164 patients between September 2011 through November 2024 were included (de-identified). As shown in the tables and figures above, the data was highly imbalanced across the different note-level metadata categories, reflecting real-world documentation patterns and nuances. Most notes were progress notes (79%), and nearly half of all notes occurred more than 180 days prior to pre-screening. Notes originated from 29 different specialties with approximately 20% coming from unknown specialties due to limitations in data capture in the Epic Clarity database.

### Modeling and Interpretation

#### Overall Model (All Exclusions)

```{r}
table(model_df$pred_overall_excl) # reminder of the distributioin --> pretty even distribution of notes with and without LLM-derived exclusion signals
```

```{r}
overall.fit <- glm(pred_overall_excl ~ note_type_final + specialty + 
                     encounter_type + window_bin, 
                  data = model_df, 
                  family = binomial())
summary(overall.fit)
# exp(cbind(OR = coef(overall.fit), CI = confint(overall.fit))) #OR and 95%CI
```

```{r}
coef_pvalues <- as.data.frame(summary(overall.fit)$coefficients) %>% arrange(`Pr(>|z|)`) %>% filter(`Pr(>|z|)` < 0.05) %>% rownames_to_column("term") %>% 
  rename(estimate = Estimate,
         standard_error = `Std. Error`,
         z_value = `z value`,
         p_value = `Pr(>|z|)`) 

or_ci <- exp(cbind(OR = coef(overall.fit), CI = confint(overall.fit))) %>% 
  as.data.frame() %>% 
  rownames_to_column("term") %>% 
  rename(odds_ratio = OR,
         ci_min = `2.5 %`,
         ci_max = `97.5 %`)

glm_overall <- coef_pvalues %>% 
  inner_join(or_ci, by='term') %>%
  filter(
    term != "(Intercept)",
    !is.na(odds_ratio),
    p_value < 0.05
  )

overall_model_odds_plot <- glm_overall %>%
  mutate(term = fct_reorder(term, odds_ratio)) %>%
  ggplot(aes(x = term, y = odds_ratio)) +
  geom_errorbar(
    aes(ymin = ci_min, ymax = ci_max),
    width = 0.2,
    size  = 0.6
  ) +
  geom_point(size = 1.8, color = osa_palette[1]) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10(breaks =scales::log_breaks(n = 10),
  labels = scales::label_number()) + # added this because of wide confidence intervals
  coord_flip() +
  labs(
    x = NULL,
    y = "Odds ratio (log10 scale)",
    title = "Predictors associated with any exclusion signal (where p < 0.05)"
  ) +
  theme_osa(base_size = 10) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.title.x = element_text(face = "bold"),
    plot.title = element_text(size = 13, face = "bold")
  )

ggsave(
  filename = "../figs/overall_model_odds_plot.png",
  width = 12,
  height = 8,
  dpi = 300,
  bg = "white"
)
```

```{r}
# table view of of odds ratio, 95% CI, p_value < 0.05
glm_overall %>% 
  mutate(across(c(estimate, standard_error, odds_ratio, ci_min, ci_max), ~round(.x, digits = 2))) %>% 
  select(term, p_value, odds_ratio, ci_min, ci_max) %>% 
  mutate(`95% CI` = 
           paste0("(", ci_min, "-", ci_max, ")")) %>% 
  distinct(term, p_value, odds_ratio, `95% CI`)
```

##### Model Interpretation

This multivariable logistic regression model identified several note-level metadata features that were significantly associated with LLM-identified exclusion signals across the three exclusion categories: clinical contraindications, recent procedures, and sleep-specific factors.

Encounter type and temporal proximity from pre-screening date were the strongest predictors. Compared to the reference encounter category, notes originating from office visits, hospital encounters, telemedicine visits, and reconciled outside data were substantially more likely to contain exclusion-relevant information. Temporal proximity to pre-screening was also a strong predictor of exclusion relevance. Notes within 0-30 days of pre-screening had over four times higher odds than other temporal bins to contain exclusion signals. These variables also showed relatively narrow confidence intervals which indicates that encounter types and timing were the most reliable predictors.

The model also showed that certain specialties had higher odds of containing exclusion-relevant context. The most notable include Renal, Neurology, Cardiology, Heme/Onc, Surgery - Other, Sleep Medicine, and PM&R specialties. However despite the results, confidence intervals were wide given sparsity in notes across the specialties. This indicates greater uncertainty in these effects.

Finally, ED Notes, H&P Notes, Operative Notes and Progress Notes were all significantly less likely than the reference standard to contain exclusion signals. Confidence intervals for note types were also wide which, suggests greater uncertainty in these effects.

In summary, these results suggests that exclusion-relevant information is more strongly associated with clinical care context and recency than with note type.

##### Bootstrapping

To check whether these feature effects were specific to this sample, I used bootstrapping to see how stable the model was when the data were resampled.

```{r}
library(boot)
```

```{r}
set.seed(123)
```

```{r}
boot_overall <- function(data, indices) {
  d <- data[indices, ]

  model <- glm(pred_overall_excl ~ note_type_final + specialty + 
                     encounter_type + window_bin, 
                  data = d, 
                  family = binomial())

  return(coef(model))
}
```

```{r}
set.seed(123)

boot_results <- boot(
  data = model_df,
  statistic = boot_overall,
  R = 200
)
```

```{r}
boot_results
```

```{r}
boot_df <- as.data.frame(boot_results$t)

coef_names <- names(coef(overall.fit))
colnames(boot_df) <- coef_names

summary_overall_boot_results <- boot_df %>%
  pivot_longer(everything(), names_to = "feature", values_to = "log_odds") %>%
  group_by(feature) %>%
  summarise(
    median_log_odds = median(log_odds, na.rm = TRUE),
    posneg_sign_pct = mean(sign(log_odds) == 
                             sign(median(median_log_odds, na.rm = TRUE))) * 100,
    median_OR = exp(median_log_odds),
    IQR_low = exp(quantile(log_odds, 0.25, na.rm = TRUE)),
    IQR_high = exp(quantile(log_odds, 0.75, na.rm = TRUE))
  ) %>%
  arrange(desc(posneg_sign_pct))

summary_overall_boot_results
```

```{r}
overall_model_bootstrap_plot <- summary_overall_boot_results %>%
  filter(!feature %in% c("specialtyPathology", "(Intercept)")) %>% 
  mutate(term = fct_reorder(feature, median_OR)) %>%
  ggplot(aes(x = term, y = median_OR, color = posneg_sign_pct)) +
  geom_errorbar(
    aes(ymin = IQR_low, ymax = IQR_high),
    width = 0.2,
    size  = 0.6   
  ) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Median Odds ratio (log scale)",
    title = "Stability of Effects Across Resampling (Any Exclusion Model)",
    subtitle = "Points = median OR, bars = bootstrap IQR",
    color = "Direction Stability (%)"
  ) +
  theme_osa(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.title.x = element_text(face = "bold"),
    plot.title = element_text(size = 13, face = "bold")
  ) +
  scale_color_gradientn(
    colors = c(
      "#7FCDBB",  
      "#2C7FB8"  
    )
  )

ggsave(
  filename = "../figs/overall_model_bootstrap_plot.png",
  width = 12,
  height = 8,
  dpi = 300,
  bg = "white"
)
```

As shown in the results above, bootstrapping demonstrated that most metadata features had consistent effect directions across re-sampled datasets. However, the large bootstrap interquartile ranges for certain note types indicate uncertainty in the strength of these effects, despite consistent directionality. Overall the results support the stability in the direction of effects in the overall model, but also highlights the variability in effect size for note types.

#### Clinical Contraindications Model

```{r}
table(model_df$pred_clinical_contra) # reminder of the distributioin --> pretty even distribution of notes with and without LLM-derived exclusion signals
```

```{r}
majority_0 <- sum(model_df$pred_clinical_contra == 0)
minority_1 <- sum(model_df$pred_clinical_contra == 1)

model_df_clinical <- model_df %>%
  mutate(class_weight = ifelse(pred_clinical_contra == 1, majority_0 / minority_1, minority_1 / majority_0))

# checking class imbalance correction with weighting
aggregate(class_weight ~ pred_clinical_contra, data = model_df_clinical, mean)
with(model_df_clinical, tapply(class_weight, pred_clinical_contra, sum))

clinical.fit <- glm(pred_clinical_contra ~ note_type_final + specialty + 
                     encounter_type + window_bin, 
                  data = model_df_clinical, 
                  family = binomial(),
    weights = class_weight)
summary(clinical.fit)
# exp(cbind(OR = coef(overall.fit), CI = confint(overall.fit))) #OR and 95%CI
```

```{r}
coef_pvalues_clinical <- summary(clinical.fit)$coefficients %>% 
  as.data.frame() %>%
  arrange(`Pr(>|z|)`) %>% filter(`Pr(>|z|)` < 0.05) %>% rownames_to_column("term") %>% 
  rename(estimate = Estimate,
         standard_error = `Std. Error`,
         z_value = `z value`,
         p_value = `Pr(>|z|)`) 

or_ci_clinical <- exp(cbind(OR = coef(clinical.fit), CI = confint(clinical.fit))) %>% 
  as.data.frame() %>% 
  rownames_to_column("term") %>% 
  rename(odds_ratio = OR,
         ci_min = `2.5 %`,
         ci_max = `97.5 %`)

glm_model_clinical <- coef_pvalues_clinical %>% 
  inner_join(or_ci, by='term') %>%
  filter(
    term != "(Intercept)",
    !is.na(odds_ratio),
    p_value < 0.05
  )

glm_model_clinical %>%
  mutate(term = fct_reorder(term, odds_ratio)) %>%
  ggplot(aes(x = term, y = odds_ratio)) +
  geom_errorbar(
    aes(ymin = ci_min, ymax = ci_max),
    width = 0.2,
    size  = 0.6
  ) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10() + # added this because of wide confidence intervals
  coord_flip() +
  labs(
    x = NULL,
    y = "Odds ratio (log10 scale)",
    title = "Predictors associated with clinical contraindications exclusion signals (where p < 0.05)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank()
  )
```

```{r}
# table view of of odds ratio, 95% CI, p_value < 0.05
glm_model_clinical %>% 
  mutate(across(c(estimate, standard_error, odds_ratio, ci_min, ci_max), ~round(.x, digits = 2))) %>% 
  select(term, p_value, odds_ratio, ci_min, ci_max) %>% 
  mutate(`95% CI` = 
           paste0("(", ci_min, "-", ci_max, ")")) %>% 
  distinct(term, p_value, odds_ratio, `95% CI`)
```

##### Model Interpretation

This logistic regression model identified note-level metadata associated with exclusion signals specific to clinical contraindications.

Encounter type remained the strongest predictor. Hospital encounters, office visits, telemedicine encounters, and reconciled outside data were strongly associated with higher odds of containing exclusion-relevant content, which is consistent with the results from the any-exclusion model.

Specialty-level effects also remained important predictors. Cardiology, Hematology/Oncology, Neurology, Psychiatry, and Renal specialties continued to show higher odds of containing exclusion-relevant information which is consistent with conditions that lead to skipping patients for recruitment. Like in the any-exclusion model, specialty-levels effects still showed relatively wide confidence intervals, indicating greater uncertainty in these effects.

Unlike the overall model, temporal proximity was no longer significant. None of the temporal bins were significantly associated with clinical contraindications. Additionally, though ED, operative, and H&P notes still had lower odds of containing exclusion-relevant content, while progress notes were no longer statistically significant predictors. Confidence intervals were also wide which again suggests greater uncertainty in these effects for these features.

Overall, the results from this model suggests that clinical contraindications are more strongly associated with clinical care context (encounter type and specialty) than with recency or note type which aligns with the expectation that chronic conditions remain relevant regardless of timing.

#### Other (Procedural & Recent Events, Sleep-Specific) Model

```{r}
table(model_df$pred_other_excl) # reminder of the distributioin --> pretty even distribution of notes with and without LLM-derived exclusion signals
```

Because the distribution of notes without and without exclusions are imbalanced (78%/22%) with more notes not having procedural & recent events and sleep-related exclusions, inverse-frequency weighting was applied.

```{r}
majority_0_other <- sum(model_df$pred_other_excl == 0)
minority_1_other <- sum(model_df$pred_other_excl == 1)

model_df_other <- model_df %>%
  mutate(class_weight = ifelse(pred_other_excl == 1, majority_0_other / minority_1_other, minority_1_other / majority_0_other))

# checking class imbalance correction with weighting
aggregate(class_weight ~ pred_other_excl, data = model_df_other, mean)
with(model_df_other, tapply(class_weight, pred_other_excl, sum))

other.fit <- glm(pred_other_excl ~ note_type_final + specialty + 
                     encounter_type + window_bin, 
                  data = model_df_other, 
                  family = binomial(),
    weights = class_weight)
summary(other.fit)
# exp(cbind(OR = coef(overall.fit), CI = confint(overall.fit))) #OR and 95%CI
```

```{r}
coef_pvalues_other <- summary(other.fit)$coefficients %>% 
  as.data.frame() %>%
  arrange(`Pr(>|z|)`) %>% filter(`Pr(>|z|)` < 0.05) %>% rownames_to_column("term") %>% 
  rename(estimate = Estimate,
         standard_error = `Std. Error`,
         z_value = `z value`,
         p_value = `Pr(>|z|)`) 

or_ci_other <- exp(cbind(OR = coef(other.fit), CI = confint(other.fit))) %>% 
  as.data.frame() %>% 
  rownames_to_column("term") %>% 
  rename(odds_ratio = OR,
         ci_min = `2.5 %`,
         ci_max = `97.5 %`)

glm_model_other <- coef_pvalues_other %>% 
  inner_join(or_ci, by='term') %>%
  filter(
    term != "(Intercept)",
    !is.na(odds_ratio),
    p_value < 0.05
  )

glm_model_other %>%
  mutate(term = fct_reorder(term, odds_ratio)) %>%
  ggplot(aes(x = term, y = odds_ratio)) +
  geom_errorbar(
    aes(ymin = ci_min, ymax = ci_max),
    width = 0.2,
    size  = 0.6
  ) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10() + # added this because of wide confidence intervals
  coord_flip() +
  labs(
    x = NULL,
    y = "Odds ratio (log10 scale)",
    title = "Predictors associated with 'Other' exclusion signals (where p < 0.05)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank()
  )
```

```{r}
# table view of of odds ratio, 95% CI, p_value < 0.05
glm_model_other %>% 
  mutate(across(c(estimate, standard_error, odds_ratio, ci_min, ci_max), ~round(.x, digits = 2))) %>% 
  select(term, p_value, odds_ratio, ci_min, ci_max) %>% 
  mutate(`95% CI` = 
           paste0("(", ci_min, "-", ci_max, ")")) %>% 
  distinct(term, p_value, odds_ratio, `95% CI`)
```

##### Model Interpretation

This logistic regression model identified note-level metadata associated with exclusion signals related to recent procedures and sleep-specific exclusions.

Encounter type and temporal proximity from pre-screening were the strongest predictors. Notes from hospital encounters, office visits, telemedicine encounters, reconciled outside data and other encounter types were associated with higher odds of containing exclusion-relevant information.

Temporal proximity to pre-screening was also a strong predictor. Notes within 0-30 days of pre-screening had more than eleven times the odds of containing exclusion-relevant information and overall were the most statistically significant signal. Though the effects were smaller, notes within 31-180 days of pre-screening were also significantly associated with exclusion content. This indicates that exclusions signals are mainly concentrated in recent notes, but older notes were still relevant.

Compared to the previous two models, specialty-level effects were limited. Only PM&R and Endocrinology were statistically significant predictors. Additionally, while ED and Operative notes were significantly less likely to contain exclusion-relevant content, other note types were not meaningful predictors. Confidence intervals for note-type effects were also wide, indicating uncertainty in these effects compared to encounter type and temporal proximity from pre-screening, which had narrow confidence intervals.

These findings show that recent procedural and sleep-related exclusions are primary driven by encounter context and recency of documentation as rather than note type or specialty. This again reinforces the important of prioritizing recent events and certain encounter types for extracting high-yield notes.

### Bootstrapping for Sensitivity Analysis

After fitting the regression models, bootstrapping was performed to assess the stability of the regression coefficients with re-sampling.

##### Any Exclusions Model

```{r}
library(boot)
```

```{r}
set.seed(123)
```

```{r}
boot_overall <- function(data, indices) {
  d <- data[indices, ]

  model <- glm(pred_overall_excl ~ note_type_final + specialty + 
                     encounter_type + window_bin, 
                  data = d, 
                  family = binomial())

  return(coef(model))
}
```

```{r}
set.seed(123)

boot_results <- boot(
  data = model_df,
  statistic = boot_overall,
  R = 1000
)
```

```{r}
boot_results
```

```{r}
boot_df <- as.data.frame(boot_results$t)

coef_names <- names(coef(overall.fit))
colnames(boot_df) <- coef_names

summary_overall_boot_results <- boot_df %>%
  pivot_longer(everything(), names_to = "feature", values_to = "log_odds") %>%
  group_by(feature) %>%
  summarise(
    median_log_odds = median(log_odds, na.rm = TRUE),
    posneg_sign_pct = mean(sign(log_odds) == 
                             sign(median(median_log_odds, na.rm = TRUE))) * 100,
    median_OR = exp(median_log_odds),
    IQR_low = exp(quantile(log_odds, 0.25, na.rm = TRUE)),
    IQR_high = exp(quantile(log_odds, 0.75, na.rm = TRUE))
  ) %>%
  arrange(desc(posneg_sign_pct))

summary_overall_boot_results
```

```{r}
summary_overall_boot_results %>%
  filter(!feature %in% c("specialtyPathology", "(Intercept)")) %>% 
  mutate(term = fct_reorder(feature, median_OR)) %>%
  ggplot(aes(x = term, y = median_OR, color = posneg_sign_pct)) +
  geom_errorbar(
    aes(ymin = IQR_low, ymax = IQR_high),
    width = 0.2,
    size  = 0.6   
  ) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Median Odds ratio (log10 scale)",
    title = "Stability of Note Metadata Effects Across Bootstrap Resampling",
    subtitle = "Dots = median odds ratio; bars = bootstrap interquartile range (IQR)",
    color = "% of Direction Stability"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank()
  ) +
  scale_color_gradientn(
    colors = c(
      "#7FCDBB",  
      "#2C7FB8"  
    )
  )
```

```{r}
summary_overall_boot_results %>% 
  filter(!feature %in% c('(Intercept)', 'specialtyPathology')) %>% 
  filter(posneg_sign_pct > 90.0) %>% summarize(n())
summary_overall_boot_results %>% 
  filter(!feature %in% c('(Intercept)', 'specialtyPathology')) %>% 
  filter(posneg_sign_pct <= 50.0) %>% 
  summarize(n())
```

As shown in the results above, bootstrapping demonstrated that most metadata terms had consistent effect directions across re-sampled datasets, with 29 terms retaining the same directional affect in more than 90% of bootstrap iterations and no features behaving randomly (\<= 50%) under boostrapping. However, the large bootstrap interquartile ranges for certain note types and Specialty Care indicate uncertainty in the strength of these effects, relatively despite consistent directionality. Overall, the results support the stability in the direction of effects in the overall model, while also highlights the variability in effect size, especially for note types.

##### Clinical Contraindications Model

```{r}
boot_clinical <- function(data, indices) {
  d <- data[indices, ]

  model <- glm(pred_clinical_contra ~ note_type_final + specialty + 
                     encounter_type + window_bin, 
                  data = d, 
                  family = binomial())

  return(coef(model))
}

```

```{r}
set.seed(123)

boot_results_clinical <- boot(
  data = model_df,
  statistic = boot_clinical,
  R = 1000
)
```

```{r}
boot_results_clinical
```

```{r}
boot_df_clinical <- as.data.frame(boot_results_clinical$t)

coef_names <- names(coef(clinical.fit))
colnames(boot_df_clinical) <- coef_names

summary_clinical_boot_results <- boot_df_clinical %>%
  pivot_longer(everything(), names_to = "feature", values_to = "log_odds") %>%
  group_by(feature) %>%
  summarise(
    median_log_odds = median(log_odds, na.rm = TRUE),
    posneg_sign_pct = mean(sign(log_odds) == 
                             sign(median(median_log_odds, na.rm = TRUE))) * 100,
    median_OR = exp(median_log_odds),
    IQR_low = exp(quantile(log_odds, 0.25, na.rm = TRUE)),
    IQR_high = exp(quantile(log_odds, 0.75, na.rm = TRUE))
  ) %>%
  arrange(desc(posneg_sign_pct))

summary_clinical_boot_results
```

```{r}
summary_clinical_boot_results %>%
  filter(!feature %in% c("specialtyPathology", "(Intercept)")) %>% 
  mutate(term = fct_reorder(feature, median_OR)) %>%
  ggplot(aes(x = term, y = median_OR, color = posneg_sign_pct)) +
  geom_errorbar(
    aes(ymin = IQR_low, ymax = IQR_high),
    width = 0.2,
    size  = 0.6   
  ) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Median Odds ratio (log10 scale)",
    title = "Stability of Note Metadata Effects Across Bootstrap Resampling (Clinical Model)",
    subtitle = "Dots = median odds ratio; bars = bootstrap interquartile range (IQR)",
    color = "% of Direction Stability"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank()
  ) +
  scale_color_gradientn(
    colors = c(
      "#7FCDBB",  
      "#2C7FB8"  
    )
  )
```

```{r}
summary_clinical_boot_results %>% 
  filter(!feature %in% c('(Intercept)', 'specialtyPathology')) %>% 
  filter(posneg_sign_pct > 90.0) %>% summarize(n())
summary_clinical_boot_results %>% 
  filter(!feature %in% c('(Intercept)', 'specialtyPathology')) %>% 
  filter(posneg_sign_pct <= 50.0) %>% summarize(n())
```

For the clinical contraindications model, bootstrapping showed that about half of the metadata features had consistent effect directions across re-sampled datasets with 20 terms retaining the same directional affect in more than 90% of bootstrap iterations and no features behaving randomly (\<= 50%) under boostrapping. Most model terms had narrow boostrap interquartile ranges indicating stable effect size with resampling. However the REI speciallity had large bootstrap interquartile indicating uncertainty in the strength of the effect size despite consistent directionality.

##### Other Exclusions Model

```{r}
boot_other <- function(data, indices) {
  d <- data[indices, ]

  model <- glm(pred_other_excl ~ note_type_final + specialty + 
                     encounter_type + window_bin, 
                  data = d, 
                  family = binomial())

  return(coef(model))
}

```

```{r}
set.seed(123)

boot_results_other <- boot(
  data = model_df,
  statistic = boot_other,
  R = 1000
)
```

```{r}
boot_results_other
```

```{r}
boot_df_other <- as.data.frame(boot_results_other$t)

coef_names <- names(coef(other.fit))
colnames(boot_df_other) <- coef_names

summary_other_boot_results <- boot_df_other %>%
  pivot_longer(everything(), names_to = "feature", values_to = "log_odds") %>%
  group_by(feature) %>%
  summarise(
    median_log_odds = median(log_odds, na.rm = TRUE),
    posneg_sign_pct = mean(sign(log_odds) == 
                             sign(median(median_log_odds, na.rm = TRUE))) * 100,
    median_OR = exp(median_log_odds),
    IQR_low = exp(quantile(log_odds, 0.25, na.rm = TRUE)),
    IQR_high = exp(quantile(log_odds, 0.75, na.rm = TRUE))
  ) %>%
  arrange(desc(posneg_sign_pct))

summary_other_boot_results
```

```{r}
summary_other_boot_results %>%
  filter(!feature %in% c("specialtyPathology", "(Intercept)")) %>% 
  mutate(term = fct_reorder(feature, median_OR)) %>%
  ggplot(aes(x = term, y = median_OR, color = posneg_sign_pct)) +
  geom_errorbar(
    aes(ymin = IQR_low, ymax = IQR_high),
    width = 0.2,
    size  = 0.6   
  ) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Median Odds ratio (log10 scale)",
    title = "Stability of Note Metadata Effects Across Bootstrap Resampling ('Other' Model)",
    subtitle = "Dots = median odds ratio; bars = bootstrap interquartile range (IQR)",
    color = "% of Direction Stability"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank()
  ) +
  scale_color_gradientn(
    colors = c(
      "#7FCDBB",  
      "#2C7FB8"  
    )
  )
```

```{r}
summary_other_boot_results %>% 
  filter(!feature %in% c('(Intercept)', 'specialtyPathology')) %>% 
  filter(posneg_sign_pct >= 90.0) %>% summarize(n())
summary_other_boot_results %>% 
  filter(!feature %in% c('(Intercept)', 'specialtyPathology')) %>% 
  filter(posneg_sign_pct <= 50.0) %>% summarize(n())
```

For the 'other' model, bootstrapping showed the most variability in the directional affects of the different terms with 18 terms retaining the same directional affect in at least 90% of bootstrap iterations. The REI specialty showed the lowest directional stability with the direction of association switching in about half of bootstrap samples. Despite this, bootstrap interquartile ranges were generally narrow for important predictors such as encounter type and temporal proximity from pre-screening, indicating stability in effect sizes across re-sampling.

**TODO**

Include a table with the following for each model

-   Strongest Effect (In terms of Odds Ratio)

-   Most Statistically Significant Result (Lowest P-Value)

-   Most Stable (is this based on smallest confidence interval?)

-   Summary (procedural recent: recent events and encounter context are more information than note type for identifying patients likely to be excluded due to procedural timing)

Describe your results and include relevant tables, plots, and code/comments used to obtain them. You may refer to the @sec-methods as needed. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

## Conclusion

This the conclusion. The @sec-results can be invoked here.
